Capturing the shape of our experiences and how we remember them

### 2018-08-17 paper/main.tex
saving progress

\section{Abstract}
{
The human memory system is adept at cataloging the rich dynamics of ongoing experience. A defining feature of our everyday experiences is that they unfold in a structured and predictable manner, governed by the laws of the physical systems that produce them. However, this temporal structure is typically absent (or severly impoverished) in traditional trial-based memory experiments. As a consequence, our understanding of how the temporal structure (i.e. the shape) of our experiences influences memorability and memory organization is limited. In this study, we investigate how people verbally recount a video by characterizing and relating the thematic dynamics, or ``trajectories,'' of the stimulus and participants' recalls. Using a novel approach based on topic modelling, we found that despite substantial variability in the quantity of information recalled across participants as well as the particular words used to describe the experience, the overall shape of the stimulus was recapitulated by most of the participants. These findings provide a window into which aspects of naturalistic experiences must be preserved, and which might be more flexible, in considering whether and how those experiences are remembered.

@@

Human memory studies typically assess memory for the ``items'' within an experience, such as the people, places and things encountered. However, real life experiences can also be described in terms of their temporal structure (i.e. how the contents relate/change over time).  Naturalistic experiences are (typically) autocorrelated in space and time on short timescales: the contents of an experience are often similar from moment to moment and change gradually. For example, many of us spend our mornings/evenings at home and much of our days at work (e.g. an office). While the specific happenings within these contexts differ, they are often highly correlated (and thus predictable) over time. Furthermore, our experiences can also be characterized by longer timescale correlations. For example, consider returning to your office after taking a lunch break. The spatial contexts before and after lunch are highly similar, but separated in time. Thus, while evaluating the ability to recall specific contents of an experience is one way to assess memory, another important dimension to consider is the temporal structure of the remembered experience, and how the temporal structure of the memory relates to the original experience. However, these recurrent and gradually changing temporal dynamics are not typically present in traditional memory studies, but are likely critical if we wish to understand how our memory systems remember our life experiences.

Despite the continuous and (often) gradually changing nature of naturalistic experiences, prior research suggests that our memories for those same experiences may be more descrete and organized around particular ``events'' \citep{Radv12, RadvZack11, BrunEtal18}. Short timescale autocorrelations in the contents of an experience are thought associatively link information in memory, while more drastic changes in perceptual or semantic aspects of an experience (i.e. ``event boundaries'') may lead to discontinuities in memory \citep{HeusEtal18, BrunEtal18, Radv12}. There is now compelling evidence that event boundaries can influence memory organization when learning sequences of items \citep{HeusEtal18, DuBrDava13}, during navigation \citep{BrunEtal18}, and narrative comprehension \citep{ZwaaRadv98, EzzyDava11}.  However, whether people naturally recount their experiences in a way that is structured around events is not known.

In this study, we analyzed an open dataset in which participants view and then verbally recount an episode of the BBC series \textit{Sherlock} \citep{ChenEtal17}. To capture the flow of the video's content (i.e. the storyline), we developed a novel computational approach based on topic modeling \citep{BleiEtal03} and hidden Markov models \citep{Rabi89, BaldEtal17}. We propose that the temporal evolution of an experience's content gives it a unique geometric ``shape'' or ``trajectory'', defined by how the contents of the experience relate and change over time. We hypothesized that when recounting naturalistic experiences, the precise details remembered and language used may differ across people, but the shape of the stimulus would be preserved.

@@

\subsubsection{A model to capture the temporal and semantic structure of naturalistic stimuli}
We fit a topic model \citep{BleiEtal03} to hand-annotated text descriptions of scenes from the video. The text descriptions contained details of the scene such as the characters, location, and a short summary of the scene (see Fig.\ref{fig:schematic} for example text, see \ref{sec:methods} for analysis details). As depicted in Fig.~\ref{fig:model}a, the topic vectors are sparse and change slowly over time. Furthermore, there are clear transitions from one topic `state' to the next, possibly indexing scene transitions in the stimulus. To get a better handle on this temporal structure, we computed a timepoints (1976) by timepoints (1976) correlation matrix of the video model (Fig.~\ref{fig:model}c).  This correlation matrix reveals that the model has a strong, block-diagonal structure with very low off-diagonal correlation values, suggesting the representations of each block are unique and highly descriminable.

\subsubsection{Modeling verbal recall}
After watching the video, participants verbally recalled (in order) as much of the episode as they could.  We used the same topic model (fit with the text descriptions of the video) to transform participants' verbal recall transcripts. An example participant's (\#13) recall model is plotted in Fig.~\ref{fig:model}b. Like the video model, topic vectors were sparse and changed gradually.  Then, we investigated the temporal structure of the recall matrices. From each participant's recall model, we computed a timepoint (68 to 294) by timepoints by timepoints (range: 68 to 294) correlation matrix. An example participant's (\#13) correlation matrix is plotted in Fig.~\ref{fig:model}d). Like the video model, each participants' recall correlation matrix exhibited a strong block-diagonal structure (Supp. Fig.~\ref{fig:corrmats})). Notably, this suggests that participants recounted the video in discriminable segments, potentially related to the recall of specific events (or scenes) from the video.

\begin{figure}[ht!]
\centering
\includegraphics[width=\textwidth]{figs/supp1_eventseg.pdf}
\caption{\small \textbf{Modelling naturalistic stimuli and recall.} A depiction of our analysis pipeline. For all plots, darker colors indicate greater values and the range of each plot is 0-1.  A). A timepoints (1976) by topics (100) matrix representing the video stimulus.  Each row represents the most likely mixture of topics for a given timepoint (i.e. topic weights). Each column represents a different topic. B). A timepoints (1976) by topics (100) matrix representing participant \#13's recall. C). A viewing-time (1976) by viewing-time (1976) correlation matrix representing the correlation of each moment of the video model with every other moment of the video model. The white boxes represent `events' recovered by a hidden Markov model. D). A recall-time (294 sentences) by recall-time (294) correlation matrix for participant \#13. E). An events (34) by topics (100) matrix where each row represents the average topic vector for each event in the video model.  F). An events (27) by topics (100) matrix where each row represents the average topic vector for each event in participant \#13's recall model. G. A recall events (27) by video events (34) correlation matrix for participant \#13. The cells with a yellow border identify the video event with the highest correlation to a given recall event. F. A group averaged recall events (34) by video events (34) correlation matrix.  The cells with yellow borders are the video event with the highest correlation to a given average recall event.}
\label{fig:model}
\end{figure}

\subsubsection{Segmenting the video and recall models into `events'}
As described above, a striking feature of both the video and recall correlation matrices is a strong, block structure along the diagonal of the matrices (see Fig.~\ref{fig:model}c,d).  We hypothesized that this structure might arise from transient stability in "theme" of the video/recall (i.e. ``events''). If true, we reasoned that specific events described by a participant could be ``matched'' to specific video events by computing the similarity between their topic vectors. In other words, the video event that is most similar (i.e. correlated) to a particular recall event is the event that the participant is most likely describing. To test this idea, we segmented the video and recall models in $k$ events (i.e. states) using a hidden Markov model ~\citep{BaldEtal17}. Our algorithm determined 34 events for the video model and a range of values (range: 8-27; mean: 15.41; SD: 5.6) for the recall models (see Methods for details on choosing an optimal $k$ value).  The events discovered for the model and a representative participant (\#13) are highlighted as yellow rectangles outlining blocks along the diagonal of the correlation matrices (Fig.~\ref{fig:model}b,d).

Next, we created a video ``event model'' by averaging together neighboring topic vectors that the were classified to be in the same event, resulting in an events (34) by topics (100) matrix (Fig.~\ref{fig:model}e).  We performed the same procedure for the recall matrices (see Fig.~\ref{fig:model}f for example). Then, we computed the correlation between video and recall event models, resulting in a video events (34) by recall events (8-27, depending on the participant) correlation matrix (Fig~\ref{fig:model}g). These matrices represent the similarity (correlation) between each video event and each recall event (for each participant). To determine which video event a particular recall event was most likely describing, we computed the index of the video event with the highest correlation to the recall event (i.e. the argmax).  This is depicted in Fig.~\ref{fig:model}g as the cells highlighted with a yellow border. Interestingly, our algorithm suggests that the example participant recalled most of the video events in order.

@@

\subsubsection{The shape of a naturalistic experience is preserved in recall}
Classic approaches to studying free recall in episodic memory (such as overall accuracy and serial position curve) focus on the quantity of information recalled. More recent metrics (such as lag-CRP, semantic and temporal clustering) go further to describe list-level memory organization. However, none of these approaches capture the rich autocorrelated structure inherent in naturalistic stimuli and associated recall of those experiences. Using our topic-based approach, in the next set of analyses we test the hypothesis that naturalistic recall entails recapitulating the trajectory of a stimulus.

\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/4_trajectory.pdf}
\caption{\small \textbf{Video and recall trajectory plots.} A). 2-dimensional embedding of the video events model. The arrows indicate the forward direction of the video events. B). 2-dimensional embedding of the average recall events model.  The colors refer to the most similar video events. The directional lines connecting the points represent the true video event order. The arrows represent the average direction of all recall event transitions (i.e. a line segment connecting two consecutive recall event vectors) that intersected a box (width=.25) centered on the origin of the arrow.}
\label{fig:trajectory}
\end{figure}

To visualize the relationship between the video and recall event models, we embedded them into a 2D space (using the UMAP dimensionality reduction algorithm, for details see \ref{sec:methods}) where the points represent video/recall events and the distance between them represents their similarity in ``topic space'' (Fig.~\ref{fig:trajectory}a).   We observed that the two models have a very similar temporal evolution and geometric structure. To further quantify this correspondence and to characterize how participants navigated through the space during recall, we analyzed the angle between successive recall events. We created a grid of evenly spaced points (.25 units) in the 2D ``topic space''. For each point on the grid, we drew a circle (radius=.25) around the point and grouped together all recall transitions that intersected the circle (across subjects). To visualize the average angle, we converted each transition angle to a unit vector and then averaged the vectors together. To assess consistency in the direction of the recall transition across participants, we performed a Rayleigh test (p$<$.001, corrected at p$<$.05 using permutation procedure described in Methods). Thus participants' recalls followed the same path as the video model (Fig.~\ref{fig:trajectory}b), suggesting the shape of the stimulus was preserved despite idiosyncratic differences between participants in their recalls.

@@

\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/5_topics.pdf}
\caption{\small \textbf{Analysis of topics by event memorability.} A). Group-averaged correlation between each video event and closest recall event (per subject). Error bars represent 95\% confidence intervals. B). Wordles (top 200 words) representing a weighted average of video event topic vectors weighted by event memorability (e.g. the correlation values in A). The top wordle (green) contains words from the most remembered events and the bottom (red) contains words from least memorable events. The word sizes are proportional to the word's ``activation''. C). Trajectory represents the video event model embedded in a 2D space using UMAP. The large colored dots are video events. The small colored dots are all individual recall events across all participants, where the colors refer to the closest video event. The circular wordles represent the top 3 most (green circles) and least (red) memorable video events.  The left side of the wordle circles contain words associated with the video event vector and the right side contains words associated with the average recall event vector.}
\label{fig:topics}
\end{figure}

@@

@@

% Supplemental materials
\begin{suppfigure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/supp2_corrmats.pdf}
\caption{\small \textbf{Recall model correlation matrices and event segmentation fits.} Each participant's timepoint-by-timepoint recall correlation matrix.  The yellow boxes represent ``events'' identified by a hidden Markov model.}
\label{fig:corrmats}
\end{suppfigure}

>>>>

\section{Abstract}
{
The human memory system is adept at cataloging the rich dynamics of ongoing experience. A defining feature of our everyday experiences is that they unfold in a structured and predictable manner. This temporal structure gives each of our experiences (and thus our memories for those experiences) a unique shape.  However, temporal structure is typically absent (or severly impoverished) in traditional trial-based memory experiments. As a consequence, our understanding of how the temporal structure (i.e. the shape) of our experiences influences memorability and memory organization is limited. In this study, we investigate how people verbally recount a video by characterizing and relating the thematic dynamics, or ``trajectories,'' of the stimulus and participants' recalls. Using a novel approach based on topic modelling, we found that despite substantial variability in the quantity of information recalled across participants as well as the particular words used to describe the experience, the overall shape of the stimulus was recapitulated by most of the participants. These findings provide a window into which aspects of naturalistic experiences must be preserved, and which might be more flexible, in considering whether and how those experiences are remembered.

@@

Human memory studies typically assess memory for the elements within an experience, such as the people, places and things encountered. However, real life experiences can also be described in terms of their temporal structure (i.e. how the contents relate/change over time).  Naturalistic experiences are (typically) autocorrelated in space and time on short timescales: the contents of an experience are often similar from moment to moment and change gradually. For example, many of us spend our mornings/evenings at home and much of our days at work (e.g. an office). While the specific happenings within these contexts differ, they are often highly correlated (and thus predictable) over time. Furthermore, our experiences can also be characterized by longer timescale correlations. For example, consider returning to your office after taking a lunch break. The spatial contexts before and after lunch are highly similar, but separated in time. Thus, while evaluating the ability to recall specific contents of an experience has been a fruitful way to learn about human memory, another important dimension to consider is the temporal structure of the remembered experience, and how the temporal structure of the memory relates to the original experience. However, these recurrent and gradually changing temporal dynamics are not typically present in traditional memory studies, but are likely critical if we wish to understand how our memory systems remember our life experiences.

While our life experiences often change gradually, they can also change rapidly (e.g. walking through a doorway) \citep{Radv12}. Prior research suggests that these sharp transitions (coined in the literature as ``event boundaries'') during an experience may serve to descretize/organize our memories into ``events'' \citep{Radv12, RadvZack11, BrunEtal18, HeusEtal18, ClewDava18}. Short timescale autocorrelations in the contents of an experience are thought associatively link information in memory, while more drastic changes in perceptual or semantic aspects of an experience may lead to discontinuities in memory \citep{HeusEtal18, BrunEtal18, Radv12}. There is now compelling evidence that event boundaries can influence memory organization when learning sequences of items \citep{HeusEtal18, DuBrDava13}, during navigation \citep{BrunEtal18}, and narrative comprehension \citep{ZwaaRadv98, EzzyDava11}.  However, whether people naturally recount their experiences in a way that is structured around events is not known.

In this study, we analyzed an open dataset in which participants view and then verbally recall an episode of the BBC series \textit{Sherlock} \citep{ChenEtal17}. To capture the flow of the video's content (i.e. the storyline), we developed a novel computational approach based on topic modeling \citep{BleiEtal03} and hidden Markov models \citep{Rabi89, BaldEtal17}. This model casts naturalistic stimuli as high-dimensional topic ``trajectories'', whereby the theme of each moment of the video is represented as a topic vector. Successful remembering entails verbally ``traversing'' the path formed by the stimulus topic trajectory as closely as possible. We hypothesized that successful memory for a naturalistic stimulus is given by a recall trajectory that 1) has temporal structure similar to the stimulus 2) closely follows the path of the stimulus trajectory and 3) covers the majority of the stimulus path.

@@

\subsubsection{Modeling the naturalistic stimulus and verbal recall}
We fit a topic model \citep{BleiEtal03} to hand-annotated text descriptions of scenes from the video. The text descriptions contained details of the scene such as the characters, location, and a short summary of the scene (see Fig.\ref{fig:schematic} for example text, see \ref{sec:methods} for analysis details). As depicted in Fig.~\ref{fig:model}a, the topic vectors are sparse and change slowly over time. Furthermore, there are clear transitions from one topic `state' to the next, possibly indexing scene transitions in the stimulus. To get a better handle on this temporal structure, we computed a timepoints (1976) by timepoints (1976) correlation matrix of the video model (Fig.~\ref{fig:model}c).  This correlation matrix reveals that the model has a strong, block-diagonal structure with very low off-diagonal correlation values, suggesting the representations of each block are unique and highly descriminable.

After watching the video, participants verbally recalled (in order) as much of the episode as they could.  We used the same topic model (fit with the text descriptions of the video) to transform participants' verbal recall transcripts. An example participant's (\#13) recall model is plotted in Fig.~\ref{fig:model}b. Like the video model, topic vectors were sparse and changed gradually.  Then, we investigated the temporal structure of the recall matrices. From each participant's recall model, we computed a timepoint (68 to 294) by timepoints by timepoints (range: 68 to 294) correlation matrix. An example participant's (\#13) correlation matrix is plotted in Fig.~\ref{fig:model}d). Like the video model, each participants' recall correlation matrix exhibited a strong block-diagonal structure (Supp. Fig.~\ref{fig:corrmats})). Notably, this suggests that participants recounted the video in discriminable segments, likely related to the recall of specific events from the video.

\begin{figure}[ht!]
\centering
\includegraphics[width=\textwidth]{figs/2_eventseg.pdf}
\caption{\small \textbf{Modelling naturalistic stimuli and recall.} A depiction of our analysis pipeline. For all plots, darker colors indicate greater values and the range of each plot is 0-1.  A). A timepoints (1976) by topics (100) matrix representing the video stimulus.  Each row represents the most likely mixture of topics for a given timepoint (i.e. topic weights). Each column represents a different topic. B). A timepoints (1976) by topics (100) matrix representing participant \#13's recall. C). A viewing-time (1976) by viewing-time (1976) correlation matrix representing the correlation of each moment of the video model with every other moment of the video model. The white boxes represent `events' recovered by a hidden Markov model. D). A recall-time (294 sentences) by recall-time (294) correlation matrix for participant \#13. E). An events (34) by topics (100) matrix where each row represents the average topic vector for each event in the video model.  F). An events (27) by topics (100) matrix where each row represents the average topic vector for each event in participant \#13's recall model. G. A recall events (27) by video events (34) correlation matrix for participant \#13. The cells with a yellow border identify the video event with the highest correlation to a given recall event. F. A group averaged recall events (34) by video events (34) correlation matrix.  The cells with yellow borders are the video event with the highest correlation to a given average recall event.}
\label{fig:model}
\end{figure}

\subsubsection{Segmenting the video and recall models into `events'}
As described above, a striking feature of both the video and recall correlation matrices is a strong, block structure along the diagonal of the matrices (see Fig.~\ref{fig:model}c,d).  We hypothesized that this structure might arise from transient stability in "theme" of the video/recall (i.e. ``events''). We reasoned that specific events described by a participant could be ``matched'' to specific video events by computing the similarity between their topic vectors. In other words, the video event that is most similar (i.e. correlated) to a particular recall event is the event that the participant is most likely describing. To test this idea, we segmented the video and recall models in $k$ events (i.e. states) using a hidden Markov model ~\citep{BaldEtal17}. Our algorithm determined 34 events for the video model and a range of values (range: 8-27; mean: 15.41; SD: 5.6) for the recall models (see Methods for details on choosing an optimal $k$ value).  The events discovered for the model and a representative participant (\#13) are highlighted as yellow rectangles outlining blocks along the diagonal of the correlation matrices (Fig.~\ref{fig:model}b,d).

Next, we created a video ``event model'' by averaging together neighboring topic vectors that the were classified to be in the same event, resulting in an events (34) by topics (100) matrix (Fig.~\ref{fig:model}e).  We performed the same procedure for the recall matrices (see Fig.~\ref{fig:model}f for example). Then, we computed the correlation between video and recall event models, resulting in a video events (34) by recall events (8-27, depending on the participant) correlation matrix (Fig~\ref{fig:model}g). These matrices represent the similarity (correlation) between each video event and each recall event (for each participant). To determine which video event a particular recall event was most likely describing, we computed the index of the video event with the highest correlation to the recall event (i.e. the argmax).  This is depicted in Fig.~\ref{fig:model}g as the cells highlighted with a yellow border. Notably, our algorithm suggests that the example participant recalled most of the video events and did so in order.

@@

\subsubsection{The shape of a naturalistic experience is preserved in recall}
Classic approaches to studying free recall in episodic memory (such as overall accuracy and serial position curve) focus on the quantity of information recalled. More recent metrics (such as lag-CRP, semantic and temporal clustering) go further to describe list-level memory organization. However, none of these approaches capture the rich autocorrelated structure inherent in naturalistic stimuli and associated recall of those experiences. In the next set of analyses, we test the hypothesis that successful recall of a naturalistic stimulus entails recovering the trajectory of a stimulus.

\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/3_trajectory.pdf}
\caption{\small \textbf{Video and recall trajectory plots.} A). 2-dimensional embedding of the video events model. The arrows indicate the forward direction of the video events. B). 2-dimensional embedding of the average recall events model.  The colors refer to the most similar video events. The directional lines connecting the points represent the true video event order. The arrows represent the average direction of all recall event transitions (i.e. a line segment connecting two consecutive recall event vectors) that intersected a box (width=.25) centered on the origin of the arrow.}
\label{fig:trajectory}
\end{figure}

To visualize the relationship between the video and recall event models, we embedded them into a 2D space (using the UMAP dimensionality reduction algorithm, for details see \ref{sec:methods}) where the points represent video/recall events and the distance between them represents their similarity/distance in ``topic space'' (Fig.~\ref{fig:trajectory}a).  We observed that visually, it appears that the two models have a very similar temporal evolution and geometric structure. To further quantify this correspondence and to characterize how participants navigated through the space during recall, we analyzed the angle between successive recall events. We created a grid of evenly spaced points (.25 units) in the 2D ``topic space''. For each point on the grid, we drew a circle (radius=.25) around the point and grouped together all recall transitions that intersected the circle (across subjects). To visualize the average angle, we converted each transition angle to a unit vector and then averaged the vectors together. To assess consistency in the direction of the recall transition across participants, we performed a Rayleigh test (p$<$.001, corrected at p$<$.05 using permutation procedure described in Methods). Thus participants' recalls followed the same path as the video model (Fig.~\ref{fig:trajectory}b), suggesting the shape of the stimulus was preserved despite idiosyncratic differences between participants in their recalls.

@@

\begin{figure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/4_topics.pdf}
\caption{\small \textbf{Analysis of topics by event memorability.} A). Group-averaged correlation between each video event and closest recall event (per subject). Error bars represent 95\% confidence intervals. B). Wordles (top 200 words) representing a weighted average of video event topic vectors weighted by event memorability (e.g. the correlation values in A). The top wordle (green) contains words from the most remembered events and the bottom (red) contains words from least memorable events. The word sizes are proportional to the word's ``activation''. C). Trajectory represents the video event model embedded in a 2D space using UMAP. The large colored dots are video events. The small colored dots are all individual recall events across all participants, where the colors refer to the closest video event. The circular wordles represent the top 3 most (green circles) and least (red) memorable video events.  The left side of the wordle circles contain words associated with the video event vector and the right side contains words associated with the average recall event vector.}
\label{fig:topics}
\end{figure}

@@

%NOTE: could be useful to add a 'where this might break down' section

@@

% Supplemental materials
\begin{suppfigure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/supp1_corrmats.pdf}
\caption{\small \textbf{Recall model correlation matrices and event segmentation fits.} Each participant's timepoint-by-timepoint recall correlation matrix.  The yellow boxes represent ``events'' identified by a hidden Markov model.}
\label{fig:corrmats}
\end{suppfigure}

\begin{suppfigure}[t!]
\centering
\includegraphics[width=1\textwidth]{figs/supp2_matchmats.pdf}
\caption{\small \textbf{Video-recall event model correlation matrices.} Each participant's video event by recall event correlation matrix.  The yellow boxes represent the maximum correlation in each column.}
\label{fig:matchmats}
\end{suppfigure}
